---
title: "The basics of wisps"
author: "Michael Barkasi"
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
options(width = 10000)
```

<h2>Simulating a Poisson process</h2>

Suppose you wanted to model some process which generates tokens of a certain type over space. For example, the cells in your body generate RNA molecules through gene transcription and rabbits spread across a forest generate offspring through reproduction. If every instance of the process (e.g., every cell, or every rabbit) in the same place $x$ produces tokens at the same rate $\lambda(x)$, then the number of tokens generated in a given region will follow a Poisson distribution $\text{Pois}(n\lambda(x))$ with rate parameter equal to $n(x)\lambda(x)$, for $n(x)$ the number of process instances in the region. As Poisson distributions are linear, it is just as well to ignore $n$ and let $\lambda(x)$ be the overall rate of token generation in the region $x$. In this case, empirical measures of $\lambda(x)$ are obtained simply by summing the tokens produced by all the process instances in $x$. 

Real cases of distributed Poisson processes rarely have instances with identical rates. For example, rabbits will reproduce at different rates depending on their age and health and cells will transcribe genes at different rates depending on their type and state. Some of this variation is due to systematic factors, but much of it is due to random noise. This noise is often gamma-distributed, such that if $\lambda$ is the mean process rate in $x$, then individual process instances will have rate $\Lambda$ drawn from a gamma distribution $\text{Gam}(\lambda, \sigma_\gamma^2)$ with expected value $\lambda$ and variance $\sigma_\gamma^2$.

Let's simulate a Poisson process with gamma-distributed noise with instances spread over a one-dimensional space. The space will be divided into 100 bins: 

```{r set_spatial_axis}
x <- seq(1, 100)
```

We will randomly distribute 5,000 process instances across the space, for an average of 50 instances per bin. The variable holding these positions will be called **bin**:

```{r distribute_instances}
set.seed(123)
n_instances <- 10000
bin <- sample(x, n_instances, replace = TRUE)
```

Now suppose that the mean rate $\lambda$ is constant across space and equals 1, while the variance $\sigma_\gamma^2$ between the instances is 2. 

```{r simulate_constant_rate}
lambda <- rep(c(1, 5, 2, 7), each = length(x)/4)
var_gamma <- 2
process_lambda <- rgamma(
  n_instances, 
  shape = lambda[bin]^2/var_gamma, 
  rate = lambda[bin]/var_gamma
  )
count <- rpois(n_instances, process_lambda)
```

These simulated process counts can be plotted as a function of space:

```{r plot_constant_rate}
countdata <- data.frame(bin, count)
ggplot2::ggplot(countdata, ggplot2::aes(x=bin, y=count)) +
  ggplot2::geom_jitter(height=0, width=0.5, alpha=0.1) +
  ggplot2::labs(
    title = "Simulated Poisson process with constant rate",
    x = "Spatial bin",
    y = "Process count"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    panel.background = ggplot2::element_rect(fill = "white", colour = NA),
    plot.background  = ggplot2::element_rect(fill = "white", colour = NA)
  )
```

```{r model}

library(wispack, quietly = TRUE)

model <- wisp(countdata, fit_only = TRUE)

```


